Insight Into Lumbar Back Pain \subtitle{What the Lumbar Spine Tells About Your Life}

### 2014-07-28 draft.tex
Wrote first draft of epidemiological background chapter

Modern cohort studies often comprise medical image data.
%%
These data are hard to analyze, since segmentation algorithms are not generally available and need to be custom-made for each body structure.
%%
Segmentation data it is usually analyzed by abstracting it into key figures, making it statistically comparable to non-image variables and retrieve correlations.
%%

@@

\section{\uppercase{Epidemiological Background}}
 \ref{VIS14}
\label{sec:EpidemiologicalBackground}
\noindent \com{Background. Epidemiological Workflow, focus on statistical resilience, image data hard to analyze due to the large amount of subjects, poor image quality and lack of methods}.
\subsection{Back Pain}
\com{Back pain one of the most common diseases in the western civilization; hard to analyze; Epidemiologists interested in the \emph{healthy} aging process}.

Risk factors for are not completely known, thus epidemiologists search for new correlations in cohort studies.

>>>>

Modern cohort studies often comprise medical image data.
%%
\com{Short summary}
%%

@@

\section{\uppercase{Epidemiological Background}}
\label{sec:EpidemiologicalBackground}
%\noindent \com{Background. Epidemiological Workflow, focus on statistical resilience, image data hard to analyze due to the large amount of subjects, poor image quality and lack of methods}.
\noindent Epidemiology is the study of dissemination, causes and results of health-related states and events.
%%
Epidemiological reasoning relies on a strict statistically driven workflow \cite{Fletcher}:
%%
\begin{itemize}
%%
	\item Physicians formulate hypotheses based on observations made in their clinical practice.
%%
	\item To assess a hypothesis, epidemiologists compile a list of variables depicting it.
%%
	\item Statistical methods, such as regression analysis, assess the association of selected variables with the investigated disease.
%%
\end{itemize}
%%
Mutually dependent variables make this analysis challenging.
%%
Many diseases, such as different cancer types, are more likely with increasing age.
%%
When for example analyzing influences of nutrition to prostate cancer, the results need to be age-normalized.
%%
Age acts as an \emph{confounder} for prostate cancer.
%%
These \emph{confounding} variables are often hard to find.
%%
Statistical correlation does not imply causation--epidemiologists need to assess the medical soundness of the statistical results.

\subsection{Lower Back Pain}
% \com{Back pain one of the most common diseases in the western civilization; hard to analyze; Epidemiologists interested in the \emph{healthy} aging process}.
%
% Risk factors for are not completely known, thus epidemiologists search for new correlations in cohort studies.
Lower back pain is one of the most frequent diseases in the western civilization.
%%
Epidemiologists assume associations between lumbar back pain and lifestyle factors.
%%
These include nutrition, sporting activities and body posture at work.
%%
The exact causes as well as particularly vulnerable risk groups are not known.
%%
Interaction between risk factors are also not known.
%%

Epidemiologists want to characterize the healthy aging process of the spine.
%%
To achieve this, they have to analyze the lumbar spine shape as well as the mentioned lifestyle factors.
%% -------------------
\subsection{Epidemiological Data}
Epidemiological data can stem from a wide range of studies.
%%
The study type depends on the condition of interest.
%%
Most common are case-control studies, analyzing one specific disease and its influences.
%%
We focus on data from large scale cohort studies.
%%
These studies aim to collect as much data as possible for each subject.
%%
As a result, these data can be analyzed regarding many diseases and conditions.
%%

Epidemiological data is heterogenous and incomplete.
%%
Women-specific question can only be acquired for female subjects, data about a disease treatment only for subjects suffering from this condition.
%%
Therefore, statistical analysis has to take missing data into account.
%%

Epidemiologists acquire data using a wide range of techniques, such as medical examinations, self reported questionnaires or genetic examinations.
%%
This yields a heterogenous information space.
%%
To compare these data, information reduction techniques are applied.
%%
For example, continuous data, such as age is often discretized into age-bins.
%%
Every information reduction can introduce a bias to the data, since it reflects an assumption about the data.
%%
Using age to divide subjects into \emph{young} and \emph{old} categories can distort statistical results.
%%
This distortion is reduced with increasing number of discretization steps.
%%

Modern cohort studies often comprise medical image data.
%%
These data are hard to analyze, since segmentation algorithms are not generally available and need to be custom-made for each body structure.
%%
Segmentation data is usually analyzed by abstracting it into key figures, such as diameters or distances.
%%
These numeric values can be compared with non-image variables to retrieve correlations.
%%
